---
title: "Task_3_Descriptive_Stats.rmd"
author: "Sean Franco"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

**Name:**   Task_3_Descriptive_Stats.rmd             
**Purpose:**  To clean, and explore Ohio's registered apprentice data.      
**Author:**   Sean Franco       
**Date Created:** 2025_07_23          
**Inputs:**   'Apprenticeship Data Dictionary.xlsx'; oh_appr_person.csv; oh_appr_program.csv             
**Outputs:**  Task_3_Descriptive_Stats.html        
**Notes:**   3/3 Interview file submission.

```{r reset-sinks, include=FALSE}
while (sink.number() > 0) sink(NULL)
```

```{r include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scientific=999, digits = 3) 
```

```{R}
#install.packages('devtools')
#devtools::install_github("physicsland/ezids")
#library(devtools)

library(ezids)

data_wd <- "C:/Users/18045/Documents/R/OSU_Project/Data"
file_wd <- "C:/Users/18045/Documents/R/OSU_Project"

setwd(file_wd)
#install.packages('pacman')
library('pacman')
pacman::p_load(ggplot2, xtable, dplyr,  readxl, gridExtra, xtable, 
               readr, car, xtable, knitr, kableExtra)

#LoadData----
appr_dict <- read_excel(file.path(data_wd, 'Apprenticeship Data Dictionary.xlsx'))
appr_person_df <- read.csv(file.path(data_wd, 'oh_appr_person.csv'))
appr_prog_df <- read.csv(file.path(data_wd, 'oh_appr_program.csv'))

```

  
```{r , message=FALSE, warning=FALSE}
#applying dictionary terms to demographic information

appr_person_df <- appr_person_df %>% 
      mutate(across(c(Race, Gender, Ethnicity), ~ na_if(., "")))%>%
      mutate(
        education = case_when(education == 1 ~ '8th grade or less', 
                                    education == 1 ~ '8th grade or less',
                                    education == 2 ~ '9th to 12th grade',
                                    education == 3 ~ 'GED',
                                    education == 4 ~ 'High school or greater',
                                    education == 5 ~ 'Unknown',
                                    education == 6 ~ 'Post-secondary or technical training' ), 
                                    
            Ethnicity = case_when(Ethnicity == 'H' ~ 'Hispanic', 
                                  Ethnicity == 'N' ~ 'Not Hispanic', 
                                  Ethnicity == 'NP' ~ 'Not Provided'), 
                                  
            apprStatus = case_when(apprStatus == 'CA' ~ 'Cancelled', 
                                  apprStatus == 'CO' ~ 'Completed', 
                                  apprStatus == 'RE' ~ 'Registered', 
                                  apprStatus == 'RI' ~ 'Suspended/Reinstated/Transferred', 
                                  apprStatus == 'SU' ~ 'Suspended/Reinstated/Transferred', 
                                  apprStatus == 'TR' ~ 'Suspended/Reinstated/Transferred'), 

            Gender = case_when(Gender == "M"~ 'Male', 
                               Gender == 'F' ~ 'Female'), 

            Disabled = case_when(Disabled  == "Y"~ 'Yes', 
                               Disabled  == 'N' ~ 'No', 
                               Disabled == 'NP' ~ "Not Provided"), 
                               
            VetStatInd = case_when(VetStatInd  == "Y"~ 'Yes', 
                               VetStatInd  == 'N' ~ 'No', 
                               VetStatInd == 'NP' ~ "Not Provided")  
                               )

```

```{r , message=FALSE, warning=FALSE}
#Binning start Ages and flagging erronous ages
appr_person_df$age_flag <- appr_person_df$AgeAtStart >= 14 & appr_person_df$AgeAtStart <= 90

appr_person_df$start_age_clean <- ifelse(appr_person_df$age_flag, appr_person_df$AgeAtStart, NA)

appr_person_df$age_bin <- cut(appr_person_df$start_age_clean,
                              breaks = c(14, 20, 30, 40, 50, 60, 70, 90),
                              right = FALSE,
                              labels = c("14–19", "20–29", "30–39", "40–49", "50–59", "60–69", "70-90"))

appr_person_df$age_bin <- factor(appr_person_df$age_bin, levels = c("14–19", "20–29", "30–39", "40–49", "50–59", "60–69", "70-90"),
  ordered = TRUE)

```


```{r , message=FALSE, warning=FALSE}
#Binning starting wages 

appr_person_df$st_wage_flag <- appr_person_df$startingWage >= 7.25 & appr_person_df$startingWage <= 40

appr_person_df$st_wage_clean <- ifelse(appr_person_df$st_wage_flag, appr_person_df$startingWage, NA)

appr_person_df$st_wage_bin <- cut(appr_person_df$startingWage, 
                              breaks = c(-Inf, 7.25, 15, 20, 30, 40, Inf), 
                              right = FALSE, 
                              labels = c("Below 7.25", "7.25–14", "15–19", "20–29", "30–39", "40+"))                      

appr_person_df$st_wage_bin <- factor(appr_person_df$st_wage_bin, levels = c("Below 7.25", "7.25–14", "15–19", "20–29", "30–39", "40+"),
  ordered = TRUE)

#Binning ending wages 

appr_person_df$end_wage_flag <- appr_person_df$exitWage >= 7.25 & appr_person_df$exitWage <= 40

appr_person_df$end_wage_clean <- ifelse(appr_person_df$end_wage_flag, appr_person_df$exitWage, NA)

appr_person_df$end_wage_bin <- cut(appr_person_df$exitWage,
                               breaks = c(-Inf, 7.25, 15, 20, 30, 40, Inf),
                               right = FALSE,
                               labels = c("Below 7.25", "7.25–14", "15–19", "20–29", "30–39", "40+"))

appr_person_df$end_wage_bin <- factor(appr_person_df$end_wage_bin, levels = c("Below 7.25", "7.25–14", "15–19", "20–29", "30–39", "40+"),
  ordered = TRUE)

```

Tables of apprentice participation demographics including Ethnicity, Veteran Status, Gender Status, Race, Disabled Status, Education, and Program Age Start Bins.

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Ethnicity) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(Ethnicity, n), y = n, fill = Ethnicity)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Ethnicity Distribution",
    x = "Ethnicity",
    y = "Count"
  ) +
  theme_minimal()
```

```{r, message = FALSE, warning=FALSE}
appr_person_df %>%
  count(Ethnicity) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>%
  select(Ethnicity, Count, Percent) %>%
  kable(
    col.names = c("Ethnicity", "Count", "Percent"),
    caption = "Ethnicity Distribution"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

Veteran Status

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(VetStatInd) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(VetStatInd, n), y = n, fill = VetStatInd)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Veteran Status Distribution",
    x = "Veteran Status",
    y = "Count"
  ) +
  theme_minimal()+ 
  scale_fill_discrete(name = "Veteran Status")
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(VetStatInd) %>%
  arrange(desc(n)) %>% 
  mutate(Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))) %>% select(VetStatInd, Count, Percent) %>%
  kable(col.names = c("Veteran Status", "Count", "Percent"), caption = "Veteran Status Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Gender) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(Gender, n), y = n, fill = Gender)) +
  geom_bar(stat = "identity") +
  
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 200000, by = 15000), 
  labels = scales::comma) +
  theme_minimal()
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Gender) %>%
  arrange(desc(n)) %>% 
  mutate(Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))) %>% select(Gender, Count, Percent) %>%
  kable(col.names = c("Gender Status", "Count", "Percent"), caption = "Gender Status Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Race 

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Race) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(Race, n), y = n, fill = Race)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Racial Distribution",
    x = "Race",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 150000, by = 15000), 
  labels = scales::comma) +
  theme_minimal()
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Race) %>%
  arrange(desc(n)) %>% 
  mutate(Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))) %>% select(Race, Count, Percent) %>%
  kable(col.names = c("Race Status", "Count", "Percent"), caption = "Race Status Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Disabled) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(Disabled, n), y = n, fill = Disabled)) +
  geom_bar(stat = "identity") +
  
  labs(
    title = "Disabled Status Distribution",
    x = "Disabled Status",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 200000, by = 15000), 
  labels = scales::comma) +
  theme_minimal() + 
  scale_fill_discrete(name = "Disabled Status")
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(Disabled) %>%
  arrange(desc(n)) %>% 
  mutate(Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))) %>% select(Disabled, Count, Percent) %>%
  kable(col.names = c("Disabled Status", "Count", "Percent"), caption = "Disabled Status Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Education

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(education) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(education, n), y = n, fill = education)) +
  geom_bar(stat = "identity") +
  coord_flip() +  
  labs(
    title = "Education Level Distribution",
    x = "Education Level",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 150000, by = 15000), 
  labels = scales::comma) +
  theme_minimal() + 
  scale_fill_discrete(name = "Education Level")
```

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(education) %>%
  arrange(desc(n)) %>% 
  mutate(Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))) %>% select(education, Count, Percent) %>%
  kable(col.names = c("Education Status", "Count", "Percent"), caption = "Education Status Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Age Bins

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(age_bin) %>%
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = age_bin, y = n, fill = age_bin)) +
  geom_bar(stat = "identity") +
  #coord_flip() +  
  labs(
    title = "Age Bins",
    x = "Age Bins",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 150000, by = 5000), 
  labels = scales::comma) +
  theme_minimal() + 
  scale_fill_discrete(name = "Age Bins")
```

```{r , message=FALSE, warning=FALSE}

appr_person_df %>%
  count(age_bin) %>%
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>%
  select(age_bin, Count, Percent) %>%
  kable(
    col.names = c("Age Bins At Start of Program", "Count", "Percent"),
    caption = "Age Bins At Start of Program  Distribution"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

Apprentice Status

```{r , message=FALSE, warning=FALSE}
appr_person_df %>%
  count(apprStatus) %>%
  arrange(desc(n)) %>% 
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>% 
  ggplot( aes(x = reorder(apprStatus, n), y = n, fill = apprStatus)) +
  geom_bar(stat = "identity") +

  labs(
    title = "Apprentice Status Distribution",
    x = "Apprentice Status",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, 150000, by = 5000), 
  labels = scales::comma) +
  theme_minimal() + 
  scale_fill_discrete(name = "Apprentice Status")
```

```{r , message=FALSE, warning=FALSE}

appr_person_df %>%
  count(apprStatus) %>%
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>%
  select(apprStatus, Count, Percent) %>%
  kable(
    col.names = c("Program Status", "Count", "Percent"),
    caption = "Program Status of Program  Distribution"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

Starting and Ending Wages 

```{r , message=FALSE, warning=FALSE}

appr_person_df %>%
  count(st_wage_bin) %>%
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>%
  select(st_wage_bin, Count, Percent) %>%
  kable(
    col.names = c("Starting Wage", "Count", "Percent"),
    caption = "Starting Wage Distribution"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

```{r , message=FALSE, warning=FALSE}

appr_person_df %>%
  count(end_wage_bin) %>%
  mutate(
    Count = scales::comma(n),
    Percent = scales::percent(n / sum(n))
  ) %>%
  select(end_wage_bin, Count, Percent) %>%
  kable(
    col.names = c("Exit Wage", "Count", "Percent"),
    caption = "Exit Wage Distribution"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```


```{r , message=FALSE, warning=FALSE}

start_tab <- table(appr_person_df$st_wage_bin)
end_tab <- table(appr_person_df$end_wage_bin)

start_df <- as.data.frame(start_tab)
colnames(start_df) <- c("Wage_Bin", "Count")
start_df$Type <- "Starting Wage"

end_df <- as.data.frame(end_tab)
colnames(end_df) <- c("Wage_Bin", "Count")
end_df$Type <- "Ending Wage"

wage_plot_df <- rbind(start_df, end_df)

ggplot(wage_plot_df, aes(x = Wage_Bin, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Starting vs. Ending Wage Distribution",
    x = "Wage Bin",
    y = "Count"
  ) +
  scale_fill_manual(values = c("Starting Wage" = "#1f77b4", "Ending Wage" = "#ff7f0e")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

Below is are boxplots of starting and ending wages from the apprentice program after removing erronous 

```r
```{r wage-boxplots, results='asis', message=FALSE, warning=FALSE, error=FALSE, echo=FALSE}

wage_type <- c(rep("Starting", sum(!is.na(appr_person_df$st_wage_clean))),
               rep("Ending", sum(!is.na(appr_person_df$end_wage_clean))))

wages <- c(na.omit(appr_person_df$st_wage_clean),
           na.omit(appr_person_df$end_wage_clean))

box_df <- data.frame(Wage_Type = factor(wage_type, levels = c("Starting", "Ending")),
                     Wage = wages)

ggplot(box_df, aes(x = Wage_Type, y = Wage, fill = Wage_Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Starting" = "#1f77b4", "Ending" = "#ff7f0e")) +
  labs(title = "Boxplot of Starting vs. Ending Wages",
       x = "Wage Type", y = "Wage ($/hr)") +
  theme_minimal()

```
Apprentice employer 


